---
title: "p8105_hw3_jw4690"
author: "Jingxi Wang"
date: "2024-10-10"
output: github_document
---
```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)
```

## Problem 1

```{r}
library(p8105.datasets)
data("ny_noaa")
```

```{r}
ny_noaa_df = ny_noaa |>
  filter(!is.na(date),
         !is.na(tmax),
         !is.na(tmin)) |>
  mutate(year = year(date),
         month = month(date),
         day = day(date)) |>
  mutate(tmin = as.numeric(tmin) / 10,
         tmax = as.numeric(tmax) / 10)
```

This dataset has 7 variables and 2595176 observations. The key variables include date, sonwfall, prcp, max temperature, and min temperature.

```{r}
# find the most commonly observed values for snowfall

snowfall_counts = ny_noaa_df |>
  count(snow, name = "n_obs")

common_snowfall = snowfall_counts |>
  filter(n_obs == max(n_obs))
```

The most commonly observed values for snowfall is `r common_snowfall$snow`. I think it's because snowfall only occurs during certain month, while for much of the year, no snow is recorded, and even during winter month, many days may still have no snowfall.

```{r}
avg_max_Jan = ny_noaa_df|>
  group_by(year) |>
  filter(month == 1) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE))

Jan_plot = avg_max_Jan |>
  ggplot(aes(x = year, y = mean_tmax)) +
  geom_point(alpha = .5) +
  labs(title = "Average max temperature in January", y = "Average max temperature(°C)")
  theme(legend.position = "none")
```

```{r}
avg_max_July = ny_noaa_df|>
  group_by(year) |>
  filter(month == 7) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE))

July_plot = avg_max_July |>
  ggplot(aes(x = year, y = mean_tmax)) +
  geom_point(alpha = .5) +
  labs(title = "Average max temperature in July", y = "Average max temperature(°C)")
  theme(legend.position = "none")
```

```{r}
Jan_plot + July_plot
```

In January, temperatures are typically lower, reflecting the winter season. The temperature distribution tends to be centered around 0°C. 
In July, temperatures tend to be higher due to the summer season. There are some instances of extremely low temperatures in July, which are considered outliers.

```{r}
tmax_tmin = ny_noaa_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(title = "tmax vs tmin", x = "min temperature(°C)", y = "max temperature°C")
```

```{r}
snowfall_df = ny_noaa_df |>
  filter(snow > 0 & snow < 100)

snowfall_plot = snowfall_df |>
  ggplot(aes(x = snow, fill = factor(year))) +
  geom_density(alpha = .3) +
  labs(title = "Distribution of snowfall by year", x = "snowfall")
```

```{r}
tmax_tmin / snowfall_plot
```

The first panel uses a hexbin plot to represent the relationship between tmax and tmin, which helps visualize the density of temperature observations.
The second panel shows the distribution of snowfall (between 0 and 100 mm) separately for each year, highlighting seasonal patterns or anomalies in snowfall over time.

## Problem 2

```{r}
accel_df = read_csv("./data/nhanes_accel.csv")

covar_df = read_csv("./data/nhanes_covar.csv", skip = 4)

covar_df = covar_df |>
  mutate(
    SEQN = as.integer(SEQN),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    age = as.integer(age),
    BMI = as.numeric(BMI),
    education = factor(education, levels = c(1, 2, 3), 
                       labels = c("Less than high school", "High school equivalent", 
                                  "More than high school"))
  ) |>
  filter(age >= 21)

merged_df = inner_join(accel_df, covar_df, by = "SEQN") |>
  drop_na(sex, age, BMI, education)
```

```{r}
education_sex_table = merged_df |>
   group_by(education, sex) |>
  summarize(count = n(), .groups = "drop") |>
  arrange(education)

education_sex_table

education_sex_plot = ggplot(merged_df, aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_wrap(. ~ education) +
  labs(title = "Age Distribution by Sex and Education Level",
       x = "Age",
       y = "Count",
       fill = "Sex")

education_sex_plot
```

The table shows the number of men and women in each education category, making it easy to compare gender distribution across different educational levels.

The plot visualizes the age distribution for men and women within each education category. It can show how age ranges might differ between genders in the same education level, or how education is distributed across age groups.

```{r}
merged_df$total_activity <- rowSums(merged_df[, 2:1441])

total_activity_plot = merged_df |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_wrap(. ~ education) +
  labs(title = "Total Activity vs Age by Sex and Education Level",
       x = "Age",
       y = "Total Activity",
       color = "Sex")

total_activity_plot
```

This plot shows how total physical activity varies by age for both men and women, with separate panels for each education level. The trend lines help illustrate potential differences between genders across different age groups and education levels. The smooth lines highlight these trends, making it easier to compare overall patterns. We can observe that the activity levels decrease with age, and there could be noticeable differences in activity between men and women, as well as between educational groups.

```{r}
minute_cols = merged_df |>
  select(starts_with("min"))

Accelerometer_df = merged_df |>
  select(SEQN, sex, education, everything()) |>
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") |>
  mutate(minute = as.numeric(sub("min", "", minute)))

Accelerometer_plot = Accelerometer_df |>
  ggplot(aes(x = minute, y = activity, color = sex)) +
  geom_line(aes(group = SEQN), alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_wrap(. ~ education) +
  labs(title = "24-Hour Activity Time Course by Education Level",
       x = "Minute of the Day",
       y = "Activity (MIMS)",
       color = "Sex")

Accelerometer_plot
```

The plot shows how activity levels fluctuate over the course of the day, broken down by education level, and will use different colors to represent the sex of the participants. Across all education levels, activity tends to increase during daylight hours, with notable peaks during the morning and afternoon, followed by a decline during the night.

## Problem 3

```{r}
Jan_2020 = read_csv("./data/citibike/Jan 2020 Citi.csv/Jan 2020 Citi.csv") |>
  mutate(year = 2020, month = "January")

Jan_2024 = read_csv("./data/citibike/Jan 2024 Citi.csv/Jan 2024 Citi.csv") |>
  mutate(year = 2024, month = "January")

July_2020 = read_csv("./data/citibike/July 2020 Citi.csv/July 2020 Citi.csv") |>
  mutate(year = 2020, month = "July")

July_2024 = read_csv("./data/citibike/July 2024 Citi.csv/July 2024 Citi.csv") |>
  mutate(year = 2020, month = "July")

combined_df= bind_rows(Jan_2020, Jan_2024, July_2020, July_2024)
```

The final combines information from January 2020, January 2024, July 2020, and July 2024. Each record represents a ride in the Citi Bike system, capturing details such as ride duration, the type of bike used, the stations involved, and the user's membership status.

```{r}
summary_table = combined_df |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n()) |>
  arrange(year, month, member_casual)

summary_table
```

There is a increase in the number of rides between 2020 and 2024, both for casual riders and Citi Bike members. The number of casual rides in July is much higher compared to January in both years, likely due to the favorable weather during summer.

