---
title: "p8105_hw3_jw4690"
author: "Jingxi Wang"
date: "2024-10-10"
output: github_document
---
```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)
library(ggridges)
```

## Problem 1

```{r}
library(p8105.datasets)
data("ny_noaa")
```

```{r}
ny_noaa_df = ny_noaa |>
  filter(!is.na(date),
         !is.na(tmax),
         !is.na(tmin)) |>
  mutate(year = year(date),
         month = month(date),
         day = day(date)) |>
  mutate(tmin = as.numeric(tmin) / 10,
         tmax = as.numeric(tmax) / 10,
         prcp = as.numeric(prcp) / 10)
```

This dataset has 7 variables and 2595176 observations. The key variables include date, sonwfall, prcp, max temperature, and min temperature.

```{r}
# find the most commonly observed values for snowfall

snowfall_counts = ny_noaa_df |>
  filter(!is.na(snow)) |>
  count(snow) |>
  arrange(desc(n))

snowfall_counts
```

The most commonly observed values for snowfall is 0 mm. I think it's because snowfall only occurs during certain month, while for much of the year, no snow is recorded. 25mm, 13mm, 51mm are also common values.

```{r}
# Make a two-panel plot showing the average max temperature in January and in July in each station across years.

ny_noaa_df |>
  group_by(id, year, month) |>
  filter(month %in% c(1, 7)) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) |>
  ggplot(aes(x = year, y = mean_tmax, group = id)) +
  geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

In January, temperatures are typically lower, reflecting the winter season. The temperature distribution tends to be centered around 0°C. 
In July, temperatures tend to be higher due to the summer season. There are some instances of extremely low temperatures in July, which are considered outliers.

```{r}
#Make a two-panel plot showing (i) tmax vs tmin; and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

tmax_tmin = ny_noaa_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(title = "tmax vs tmin", x = "min temperature(°C)", y = "max temperature°C")
```

```{r}
snowfall_df = ny_noaa_df |>
  filter(snow > 0 & snow < 100)

snowfall_plot = snowfall_df |>
  ggplot(aes(x = snow, y = as.factor(year))) +
  geom_density_ridges() +
  labs(title = "Distribution of snowfall by year", x = "snowfall")
```

```{r}
tmax_tmin + snowfall_plot
```

The first panel uses a hexbin plot to represent the relationship between tmax and tmin, which helps visualize the density of temperature observations.
The second panel shows the distribution of snowfall (between 0 and 100 mm) separately for each year, highlighting seasonal patterns or anomalies in snowfall over time.

## Problem 2

```{r}
# Load, tidy, merge, and otherwise organize the data sets

accel_df = read_csv("./data/nhanes_accel.csv")

covar_df = read_csv("./data/nhanes_covar.csv", skip = 4)

covar_df = covar_df |>
  mutate(
    SEQN = as.integer(SEQN),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    age = as.integer(age),
    BMI = as.numeric(BMI),
    education = factor(education, levels = c(1, 2, 3), 
                       labels = c("Less than high school", "High school equivalent", 
                                  "More than high school"))
  ) |>
  filter(age >= 21)

merged_df = inner_join(accel_df, covar_df, by = "SEQN") |>
  drop_na(sex, age, BMI, education)
```

```{r}
# reader-friendly table for the number of men and women in each education category

education_sex_table = merged_df |>
   group_by(education, sex) |>
  summarize(count = n()) |>
  pivot_wider(names_from = sex,
              values_from = count) |>
  rename("Education Category" = education)

education_sex_table |> 
  knitr::kable(caption = "Sex and Education Category")
```

```{r}
# visualization of the age distributions for men and women in each education category

education_sex_plot = ggplot(merged_df, aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_wrap(. ~ education) +
  labs(title = "Age Distribution by Sex and Education Level",
       x = "Age",
       y = "Count",
       fill = "Sex")

education_sex_plot
```

The table shows the number of men and women in each education category, making it easy to compare gender distribution across different educational levels. We can observe that the number of men and women in each category are pretty similar.

The plot visualizes the age distribution for men and women within each education category. It can show how age ranges might differ between genders in the same education level, or how education is distributed across age groups. For those with "Less than high school," the age distribution appears fairly evenly spread across different age groups, with no clear peak for either males or females. In the "High school equivalent" group, the age distribution is somewhat more scattered, but still fairly balanced between the sexes, without a clear age range standing out. The "More than high school" group shows a larger count of people in the 30-50 age range, especially among males. 

```{r}
# plot  total activity over the day

merged_df$total_activity <- rowSums(merged_df[, 2:1441])

total_activity_plot = merged_df |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  facet_wrap(. ~ education) +
  labs(title = "Total Activity vs Age by Sex and Education Level",
       x = "Age",
       y = "Total Activity",
       color = "Sex")

total_activity_plot
```

In the "Less than high school" group, activity drops significantly after age 60 for both sexes. The "High school equivalent" group shows more fluctuation, with activity peaking around ages 40–50 and declining afterward, while females maintain slightly higher activity in older ages. In the "More than high school" group, both sexes show more stable activity levels, with females maintaining steady activity longer than males. Higher education seems to be linked with a more gradual decline in activity, especially for women.

```{r}
# Make a three-panel plot that shows the 24-hour activity time courses for each education level 

minute_cols = merged_df |>
  select(starts_with("min"))

Accelerometer_df = merged_df |>
  select(SEQN, sex, education, everything()) |>
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") |>
  mutate(minute = as.numeric(sub("min", "", minute)))

Accelerometer_plot = Accelerometer_df |>
  ggplot(aes(x = minute, y = activity, color = sex)) +
  geom_line(aes(group = SEQN), alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_wrap(. ~ education) +
  labs(title = "24-Hour Activity Time Course by Education Level",
       x = "Minute of the Day",
       y = "Activity",
       color = "Sex") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Accelerometer_plot
```

This chart shows daily activity patterns over a 24-hour period for males and females across three education levels. In all categories, activity rises sharply in the morning, peaks, and then gradually declines throughout the day, with smaller activity spikes toward the evening. The activity trends for both males and females are similar, though females tend to have slightly more consistent activity levels across the day. There are no significant differences between education levels, though the "More than high school" group shows slightly higher peaks of activity during the day.

## Problem 3

```{r}
# Import, clean, and tidy these data

Jan_2020 = read_csv("./data/citibike/Jan 2020 Citi.csv/Jan 2020 Citi.csv") |>
  mutate(year = 2020, month = "January")

Jan_2024 = read_csv("./data/citibike/Jan 2024 Citi.csv/Jan 2024 Citi.csv") |>
  mutate(year = 2024, month = "January")

July_2020 = read_csv("./data/citibike/July 2020 Citi.csv/July 2020 Citi.csv") |>
  mutate(year = 2020, month = "July")

July_2024 = read_csv("./data/citibike/July 2024 Citi.csv/July 2024 Citi.csv") |>
  mutate(year = 2024, month = "July")

combined_df= bind_rows(Jan_2020, Jan_2024, July_2020, July_2024)
```

The final dataset combines information from January 2020, January 2024, July 2020, and July 2024. Each record represents a ride in the Citi Bike system, capturing details such as ride duration, the type of bike used, the stations involved, and the user's membership status.

```{r}
summary_table = combined_df |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n()) |>
  arrange(year, month) |>
  pivot_wider(names_from = member_casual,
              values_from = total_rides) |>
  rename("Causal Riders" = casual,
         "Member Riders" = member)

summary_table |>
  knitr::kable(caption = "Total Number of Rides")
```

The number of Casual Riders has increased over time, particularly in July 2024, with a sharp rise from previous years. The number of Member Riders also shows growth, especially in July, where the figures have more than doubled from 2020 to 2024. There seems to be a seasonal pattern, where both rider types tend to increase significantly in July compared to January.

```{r}
# Make a table showing the 5 most popular starting stations for July 2024

top_starting_stations = July_2024 |>
  group_by(start_station_name) |>
  summarise(total_rides = n()) |>
  arrange(desc(total_rides)) |>
  slice(1:5)

top_starting_stations |>
  knitr::kable(caption = "The 5 Most Popular Starting Stations for July 2024")
```

The 5 most popular starting stations in July 2024 are: Pier 61 at Chelsea Piers, University Pl & E 14 St, W 21 St & 6 Ave, West St & Chambers St, and W 31 St & 7 Ave.

```{r}
# Make a plot to investigate the effects of day of the week, month, and year on median ride duration

median_duration_df = combined_df |>
  group_by(year, month, weekdays) |>
  summarize(median_duration = median(duration, na.rm = TRUE)) |>
  mutate(weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

median_duration_plot = median_duration_df |>
  ggplot(aes(x = weekdays, y = median_duration, color = month)) +
  geom_line(aes(group = interaction(year, month))) +
  facet_wrap(. ~ year) + 
  labs(title = "Median Ride Duration by Day of the Week, Month, and Year",
       x = "Day of the Week",
       y = "Median Ride Duration (minutes)",
       color = "Month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

median_duration_plot
```

This plot shows that July has consistently longer median ride durations compared to January in both 2020 and 2024, indicating a seasonal effect. In July, rides peak on Saturdays, especially in 2020, while January has more stable and shorter ride durations throughout the week. Overall, summer months (July) see more variability and longer rides, particularly on weekends, whereas winter months (January) show little variation in ride duration across the week.

```{r}
df_2024 = combined_df |>
  filter(year == 2024)

figure_2024 = df_2024 |>
  ggplot(aes(x = duration, fill = rideable_type)) +
  geom_density(alpha = .5) +
  facet_grid(member_casual ~ month) +  
  labs(title = "Impact of Month, Membership Status, and Bike Type on Ride Duration in 2024",
       x = "Ride Duration (minutes)",
       y = "Density",
       fill = "Bike Type")

figure_2024
```

The distribution of ride durations varies significantly between electric and classic bikes. Electric bikes generally have shorter ride durations, especially for casual riders. Members tend to take longer rides compared to casual riders, and there are differences between January and July, with longer rides in July. 